SSA based global code motion of identical computations

### 2017-04-14 hoist.tex
ranking and sinkability

\section{Code hoisting policies}
The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm, however, that analysis is beyond
the scope of this paper.

\subsection{Hoisting bottom up with optimistic approach}
\label{subsec:optimistic}
In this approach, the goal is to maximize the total number of hoistings in the
entire function.  This algorithm is very useful when optimizing for code-size.
We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. After that we find the common
dominator dominating all such identical computations and perform legality
checks, as described in Section~\ref{subsec:legality}. Often times it is not
possible to hoist all the instructions to one common dominator, due to legality
constraints, e.g., intersecting side-effects, or profitability constraints, e.g.,
hoisting point too far. In those cases, this algorithmn would partition the list
of identical instructions into subsets which can be partially hoisted to their
respective common dominators. The partition algorithm is described as follows:

@@

\section{Implementation Details}
\label{sec:implementation-details}

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146). All
the experiments were conducted on x86\_64 Ubuntu-Linux machine and at -Ofast
optimization level.  The results for code-hoisting are listed in
Table~\ref{tab:hoist-results}. The table lists the number of scalars, loads,
stores and calls hoisted as well as removed. For each category, the number of
instructions removed is greater or equal to the number of instructions hoisted
because each hoisting is performed only when at least one identical computation
is found.

>>>>

\section{SSA based global code motion of identical computations}
\label{sec:implementation-details}
% Ranking of instructions from Zadeck
% Liveness analysis from Ramakrishna

The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm. We have implemented a
optimistic global code motion of identical computation which uses the liveness
analysis as illustrated in Das \cite{das2012} and ranking expressions explained
in Rosen \cite{rosen1988global}.

\subsection{Optimistic code motion algorithm}
\label{subsec:optimistic}
We collect the \GVN{} of all the instructions in the function and iterate on the
list of instructions having identical \GVN{}s. The algorithm prefers hoisting to
sinking. So first we find the common dominator dominating all such identical
computations and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure. In those cases, this algorithmn would partition the list of
identical instructions into subsets which can be partially hoisted to their
respective common dominators. However, it should be noted that we only hoist
very busy expressions to avoid checking for undefined behaviors resulting because
of introducing extra computation in an execution path.

Code hoisting opens new opportunities for other hoistable candidates which were
of higher rank (depended on candidates which got hoisted). Ideally we could
iterate on lower ranking expressions first and then proceed to higher ranking
expressions in the same iteration but LLVM's GVN infrastucture does not compute
equivalence classes in a effective way. We found it simpler to just recompute
the value numbers and start finding hoistable candidates again.

For sinking however, higher ranked expressions would be sunk first. And it would
be illegal to sink higher ranked identical expressions if they are not
anticipable in the common post-dominator. For example:

\begin{verbatim}
B0: i0 = load B
B1: i1 = load A
    c1 = i1 + 10
    d1 = i0 + 20
    goto B3
B2: i2 = load A
    c2 = i2 + 10
    d2 = i0 + 20
    goto B3

B3: phi(c1, c2)
    phi(d1, d2)
\end{verbatim}

In this example (c1, c2) or (d1, d2) could be sunk first. Since (c1, c2) depend
on i1 and i2 respectively which are also in their original basic blocks, c1 and
c2 are not anticipable in B3. So without knowing the sinkability of `i1' and
`i2' it would be illegal to sink (c1, c2) to B3. On the other hand (d1, d2) can
safely be sunk because their operands are readily available at the sink point
i.e., B3. It should also be noted that, just because the expressions are
identical and operands are available it still requires a unique post-dominator
PHI to use the exact same values to be legally sinkable.

The partition algorithm is described as follows:
% TODO: Explain why all the identical computation cannot be hoisted.

@@

@@

\section{Experimental Evaluation}
\label{sec:experimental-results}
% Experimental results: Register spills, compile time, run time.
We ran \LLVMTestSuite{} (trunk:d87471f8) with the patch (trunk:86940146). All
the experiments were conducted on x86\_64 Ubuntu-Linux machine and at -Ofast
optimization level.  The results for code-hoisting are listed in
Table~\ref{tab:hoist-results}. The table lists the number of scalars, loads,
stores and calls hoisted as well as removed. For each category, the number of
instructions removed is greater or equal to the number of instructions hoisted
because each hoisting is performed only when at least one identical computation
is found.
